Optimal synaptic strategies for different timescales of memory

### 2016-03-09 Paper/SynapsePaper.tex
started pnas paper

% -*- TeX -*- -*- US -*-
\documentclass{article} % For LaTeX2e
\usepackage{nips13submit_e,times}
\input{sl_preamble.tex}
\input{sl_graphics_preamble.tex}
\graphicspath{{"Figs/"}}
% >> Only for drafts! <<
%\usepackage[notref,notcite]{showkeys}
% ----------------------------------------------------------------
%\numberwithin{equation}{section}
%\renewcommand{\baselinestretch}{1.5}
% ----------------------------------------------------------------
% New commands etc.
\input{sl_definitions.tex}
\input{sl_symbols.tex}
%
%additional symbols:
%
\DeclareMathOperator{\SNR}{SNR}
\DeclareMathOperator{\snr}{SNR}
\newcommand{\wv}{\vec{w}}
\newcommand{\wvi}{\vec{w}_\text{ideal}}
%matrices
\newcommand{\inv}{^{-1}}
\newcommand{\dg}{^\mathrm{dg}}
\newcommand{\trans}{^\mathrm{T}}
\newcommand{\I}{\mathbf{I}}
%vec of ones
\newcommand{\onev}{\mathbf{e}}
%mat of ones
\newcommand{\onem}{\mathbf{E}}
%Markov matrix
\newcommand{\MM}{\mathbf{Q}}
%prob distributions
\newcommand{\pr}{\mathbf{p}}
\newcommand{\eq}{\pr^\infty}
%first passage times
\newcommand{\fpt}{\mathbf{T}}
%off-diag first passage times
\newcommand{\fptb}{\overline{\fpt}}
%fundamental matrix
\newcommand{\fund}{\mathbf{Z}}
%other symbols for matrices
\newcommand{\Pb}{\mathbf{P}}
\newcommand{\D}{\mathbf{D}}
\newcommand{\pib}{\boldsymbol{\pi}}
\newcommand{\Lb}{\boldsymbol{\Lambda}}
\newcommand{\w}{\mathbf{w}}
\newcommand{\W}{\mathbf{W}}
\newcommand{\M}{\mathbf{M}}
\newcommand{\enc}{\mathbf{q}}
\newcommand{\frg}{\W^{\mathrm{F}}}
\newcommand{\F}{\boldsymbol{\Phi}}
%superscripts
\newcommand{\pot}{^{\text{pot}}}
\newcommand{\dep}{^{\text{dep}}}
\newcommand{\potdep}{^{\text{pot/dep}}}
%sets
\newcommand{\CS}{\mathcal{S}}
\newcommand{\CA}{\mathcal{A}}
\newcommand{\CB}{\mathcal{B}}
\newcommand{\comp}{^\mathrm{c}}
%eigenmodes
\newcommand{\uv}{\mathbf{u}}
\newcommand{\vv}{\mathbf{v}}
\newcommand{\CI}{\mathcal{I}}

%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Title info:
\title{A memory frontier for complex synapses}
%
% Author List:
%
\author{Subhaneil Lahiri and Surya Ganguli\\
Department of Applied Physics, Stanford University, Stanford CA\\
\emaillink{sulahiri@stanford.edu}, \emaillink{sganguli@stanford.edu}
%
}

\nipsfinalcopy % Uncomment for camera-ready version

@@

@@

\begin{abstract}
  An incredible gulf separates theoretical models of synapses, often described solely by a single scalar value denoting the size of a postsynaptic potential, from the immense complexity of molecular signaling pathways underlying real synapses.
  To understand the functional contribution of such molecular complexity to learning and memory, it is essential to expand our theoretical conception of a synapse from a single scalar to an entire dynamical system with many internal molecular functional states.
  Moreover, theoretical considerations alone demand such an expansion; network models with scalar synapses assuming finite numbers of distinguishable synaptic strengths have strikingly limited memory capacity.
  This raises the fundamental question, how does synaptic complexity give rise to memory?
  To address this, we develop new mathematical theorems elucidating the relationship between the structural organization and memory properties of complex synapses that are themselves molecular networks.
  Moreover, in proving such theorems, we uncover a framework, based on first passage time theory, to impose an order on the internal states of complex synaptic models, thereby simplifying the relationship between synaptic structure and function.

%Overall, we uncover general design principles governing the functional organization of complex molecular networks, and suggest new experimental observables in synaptic physiology, based on first %passage time theory, that connect molecular complexity to memory.
\end{abstract}

% references to add:
%
% Hopfield: original paper
% Somp: memory capacity of Hopfield
% Gardner: memory capacity of perceptron
% Madison/Montgomery:  State-dependent heterogeneity in synaptic depression between pyramidal cell pairs. Neuron.
% Abraham/Bear: Metaplasticity: the plasticity of synaptic plasticity.
% Grant:  Evolution of synapse complexity and diversity (Ann Rev)

\section{Introduction}\label{sec:intro}

It is widely thought that our very ability to remember the past over long time scales depends crucially on our ability to modify synapses in our brain in an experience dependent manner.
Classical models of synaptic plasticity model synaptic efficacy as an analog scalar value, denoting the size of a post-synaptic potential injected into one neuron from another.
Theoretical work has shown that such models have a reasonable, extensive memory capacity, in which the number of long term associations that can be stored by a neuron is proportional its number of afferent synapses \cite{Hopfield1982model,Amit1985hopfield,Gardner1988perceptron}.
However, recent experimental work has shown that many synapses are more digital than analog; they cannot robustly assume an infinite continuum of analog values, but rather can only take on a finite number of distinguishable strengths, a number than can be as small as two \cite{Bliss1993LTP,Petersen1998allornone,O'Connor2005switch} (though see \cite{Enoki2009graded}).
This one simple modification leads to a catastrophe in memory capacity:  classical models with digital synapses, when operating in a palimpset mode in which the ongoing storage of new memories can overwrite previous memories, have a memory capacity proportional to the logarithm of the number of synapses \cite{amit1992constraints,amit1994learning}.
Intuitively, when synapses are digital, the storage of a new memory can flip a population of synaptic switches, thereby rapidly erasing previous memories stored in the same synaptic population. This result indicates that the dominant theoretical basis for the storage of long term memories in modifiable synaptic switches is flawed.

@@

Theoretical studies of complex, metaplastic synapses have focused on analyzing the memory performance of a limited number of very specific molecular dynamical systems, characterized by a number of internal states in which potentiation and depression each induce a specific set of allowable transitions between states (\eg see \autoref{fig:models} below).
While these models can vastly outperform simple binary synaptic switches, these analyses leave open several deep and important questions.
For example, how does the structure of a synaptic dynamical system determine its memory performance?  What are the fundamental limits of memory performance over the space of all possible synaptic dynamical systems?
What is the structural organization of synaptic dynamical systems that achieve these limits?
Moreover, from an experimental perspective, it is unlikely that all synapses can be described by a single canonical synaptic model;  just like the case of neurons, there is an incredible diversity of molecular networks underlying synapses both across species and across brain regions within a single organism \cite{Emes2012synapserev}.
In order to elucidate the functional contribution of this diverse molecular complexity to learning and memory, it is essential to move beyond the analysis of specific models and instead develop a general theory of learning and memory for complex synapses.
Moreover, such a general theory of complex synapses could aid in development of novel artificial memory storage devices.  % Sutskever, memory for learning?

@@

Much of the previous work on these types of complex synaptic models has focused on understanding the memory curves of specific models, or choices of $\M\potdep$.
Two examples of these models are shown in \autoref{fig:models}. We see that they have different memory properties.
The serial model performs relatively well at one particular timescale, but it performs poorly at other times.
The cascade model does not perform quite as well at that time, but it maintains its performance over a wider range of timescales.

\begin{figure}[tbp]
 \begin{center}
 \begin{myenuma}
  \item\aligntop{\includegraphics[width=0.25\linewidth]{cascade.svg}}\label{fig:cascade}
  \item\aligntop{\includegraphics[width=0.25\linewidth]{multistate.svg}}\label{fig:multistate}
  \item\aligntop{\includegraphics[width=0.3\linewidth]{cascms.eps}}\label{fig:cascms}
 \end{myenuma}
 \end{center}
  \caption{Models of complex synapses.
  (\ref{fig:cascade}) The cascade model of \cite{Fusi2005cascade}, showing transitions between states of high/low synaptic weight (red/blue circles) due to potentiation/depression (solid red/dashed blue arrows).
  (\ref{fig:multistate}) The serial model of \cite{Leibold2008serial}.
  (\ref{fig:cascms}) The memory curves of these two models, showing the decay of the signal-to-noise ratio (to be defined in \autoref{sec:setup}) as subsequent memories are stored.
  }\label{fig:models}
\end{figure}

@@

In the previous section, in \eqref{eq:SNRcalc} we have described an analytic expression for a memory curve as a function of the structure of a synaptic dynamical system, described by the pair of stochastic transition matrices  $\M\potdep$.
Since the performance measure for memory is an entire memory curve, and not just a single number, there is no universal scalar notion of optimal memory in the space of synaptic dynamical systems.
Instead there are tradeoffs between storing proximal and distal memories;  often attempts to increase memory at late (early) times by changing $\M\potdep$, incurs a performance loss in memory at early (late) times in specific models considered so far \cite{Fusi2005cascade,Fusi2007multistate,Leibold2008serial}.
Thus our end goal, achieved in \autoref{sec:env}, is to derive an envelope memory curve in the SNR-time plane, or a curve that forms an upper-bound on the {\it entire} memory curve of {\it any} model.
In order to achieve this goal, in this section, we must first derive upper bounds, over the space of all possible synaptic models, on two different scalar functions of the memory curve: its initial SNR, and the area under the memory curve.
In the process of upper-bounding the area, we will develop an essential framework to organize the structure of synaptic dynamical systems based on first passage time theory.

@@

To maximize this flux, potentiation from a weak state must be guaranteed to end in a strong state, and depression must do the reverse.
An example of such a model is shown in \autoref{fig:max_init}(\ref{fig:max_init_pot},\ref{fig:max_init_dep}).
These models have a property known as ``lumpability'' (see \cite[\S6.3]{kemeny1960finite} for the discrete time version and \cite{burke1958markovian,Ball1993Lumpability} for continuous time).
They are completely equivalent (\ie have the same memory curve) as a two state model with transition probabilities equal to 1, as shown in \autoref{fig:max_init}(\ref{fig:binary_det}).

@@

With the states in the order described above, we can find perturbations of $\M\potdep$ that will always increase the area, whilst leaving the equilibrium distribution, $\eq$, unchanged.
Some of these perturbations are shown in \autoref{fig:perts}, see supplementary material for details.
For example, in \autoref{fig:perts}(\ref{fig:tri_right}), for two states $i$ on the left and $j$ on the right, with $j$ being more ``potentiated" than $i$ (i.e. $\smash{\eta^+_i > \eta^+_j}$), we have proven that increasing $\smash{\M\pot_{ij}}$ and decreasing $\smash{\M\dep_{ij}}$ leads to an increase in area.
The only thing that can prevent these perturbations from increasing the area is when they require the decrease of a matrix element that has already been set to 0.
This determines the topology (non-zero transition probabilities) of the model with maximal area.
It is of the form shown in \autoref{fig:env}(\ref{fig:sticky}),%\autoref{fig:max_area},
with potentiation moving one step to the right and depression moving one step to the left.
Any other topology would allow some class of perturbations (\eg\ in \autoref{fig:perts}) to further increase the area.

\begin{figure}[tbp]
 \begin{center}
 \begin{myenuma}
  \item\aligntop{\includegraphics[width=0.07\linewidth]{triangular_right.svg}}\label{fig:tri_right}\hp
  \item\aligntop{\includegraphics[width=0.07\linewidth]{triangular_left.svg}}\label{fig:tri_left}\hp
  \item\aligntop{\includegraphics[width=0.14\linewidth]{shortcut.svg}}\label{fig:shortcut}
 \end{myenuma}
 \end{center}
  \caption{Perturbations that increase the area.
  (\ref{fig:tri_right}) Perturbations that increase elements of $\M\pot$ above the diagonal and decrease the corresponding elements of $\M\dep$. It can no longer be used when $\M\dep$ is lower triangular, \ie depression must move synapses to ``more depressed'' states.
  (\ref{fig:tri_left}) Perturbations that decrease elements of $\M\pot$ below the diagonal and increase the corresponding elements of $\M\dep$. It can no longer be used when $\M\pot$ is upper triangular, \ie potentiation must move synapses to ``more potentiated'' states.
  (\ref{fig:shortcut}) Perturbation that decreases ``shortcut'' transitions and increases the bypassed ``direct'' transitions. It can no longer be used when there are only nearest-neighbor ``direct'' transitions.
  }\label{fig:perts}
\end{figure}

As these perturbations do not change the equilibrium distribution, this means that the area of \emph{any} model is bounded by that of a linear chain with the same equilibrium distribution.
The area of a linear chain model can be expressed directly in terms of its equilibrium state distribution, $\eq$, yielding the following upper bound on the area of any model with the same $\eq$ (see supplementary material):
%
\begin{equation}\label{eq:multistatearea}
  A \leq \frac{2\sqrt{N}}{r} \sum_k \brk{k - \sum_j j\eq_j} \eq_k \w_k
    = \frac{2\sqrt{N}}{r} \sum_k \abs{k - \sum_j j\eq_j} \eq_k ,
\end{equation}
%
where we chose $\w_k=\sgn[{k - \sum_j j\eq_j}]$.
We can then maximize this by pushing all of the equilibrium distribution symmetrically to the two end states.
This can be done by reducing the transition probabilities out of these states, as in \autoref{fig:env}(\ref{fig:sticky}). %\autoref{fig:max_area}.
This makes it very difficult to exit these states once they have been entered.
The resulting area is
%
\begin{equation}\label{eq:max_area}
  A \leq \frac{\sqrt{N}(M-1)}{r}.
\end{equation}
%
This analytical result is similar to a numerical result found in \cite{Barrett2008discrete} under a slightly different information theoretic measure of memory performance.

@@

Now we can pick some fixed time, $\smash{t_0}$, and maximize the SNR at that time \wrt the parameters $\smash{\brc{\CI_a,\tau_a}}$, subject to the constraints above.
This always results in a single nonzero $\smash{\CI_a}$; in essence, optimizing memory at a single time requires a single exponential.  The resulting optimal memory curve, along with the achieved memory at the chosen time, depends on $t_0$ as follows:
%
\begin{equation}\label{eq:env}
\begin{aligned}
  t_0 &\leq \frac{M-1}{r} &
   \implies
  \snr(t) &= \sqrt{N}\e^{-rt/(M-1)} &
   \implies
  \snr(t_0) &= \sqrt{N}\e^{-rt_0/(M-1)},\\
  t_0 &\geq \frac{M-1}{r} &
   \implies
  \snr(t) &= \frac{\sqrt{N}(M-1)\e^{-t/t_0}}{rt_0} &
   \implies
  \snr(t_0) &= \frac{\sqrt{N}(M-1)}{\e rt_0}.
\end{aligned}
\end{equation}
%
Both the initial SNR bound and the area bound are saturated at early times.
At late times, only the area bound is saturated.
The function $\snr(t_0)$, the green curve in \autoref{fig:env}(\ref{fig:envplot}), above forms a memory curve envelope with late-time power-law decay $\sim \smash{t_0\inv}$.
No synaptic model can have an SNR that is greater than this at any time.
We can use this to find an upper bound on the memory lifetime, $\tau(\epsilon)$, by finding the point at which the envelope crosses $\epsilon$:
%
\begin{equation}\label{eq:lifebound}
  \tau(\epsilon) \leq \frac{\sqrt{N}(M-1)}{\epsilon\e r},
\end{equation}
%
where we assume $N>(\epsilon\e)^2$.
Intriguingly, both the lifetime and memory envelope expand linearly with the number of internal states $M$, and increase as the square root of the number of synapses $N$.

This leaves the question of whether this bound is achievable.
At any time, can we find a model whose memory curve touches the envelope?
The red curves in \autoref{fig:env}(\ref{fig:envplot}) show the closest we have come to the envelope with actual models, by repeated numerical optimization of $\SNR(t_0)$ over $\M\potdep$ with random initialization and by hand designed models.

We see that at early, but not late times, there is a gap between the upper bound that we can prove and what we can achieve with actual models.
There may be other models we haven't found that could beat the ones we have, and come closer to our proven envelope.
However, we suspect that the area constraint is not the bottleneck for optimizing memory at times less than  $\CO(\frac{M}{r})$.
We believe there is some other constraint that prevents models from approaching the envelope, and currently are exploring several mathematical conjectures for the precise form of this constraint in order to obtain a potentially tighter envelope.
Nevertheless, we have proven rigorously that no model's memory curve can ever exceed this envelope, and that it is at least tight for late times, longer than $\CO(\frac{M}{r})$, where models of the form in \autoref{fig:env}(\ref{fig:sticky})%\autoref{fig:max_area}
 can come close to the envelope.

>>>>

%% PNAStwoS.tex
%% For two column PNAS articles
%% Version: Apr 15, 2008

%% BASIC CLASS FILE
\documentclass{pnastwo}

%% ADDITIONAL OPTIONAL STYLE FILES
%\usepackage{graphicx}
\input{sl_graphics_preamble.tex}
\graphicspath{{"Figs/"}}
%\usepackage{pnastwoF}
\usepackage{amssymb,amsfonts}
%\usepackage[centertags]{amsmath}
%\usepackage{hyperref}

%% OPTIONAL MACRO DEFINITIONS
\input{sl_definitions.tex}
%\input{sl_symbols.tex}
%
%additional symbols:
\newcommand{\half}{\frac{1}{2}}
\newcommand{\CO}{\mathcal{O}}
\newcommand{\dt}{\dr t}
%
\DeclareMathOperator{\SNR}{SNR}
\DeclareMathOperator{\snr}{SNR}
\newcommand{\wv}{\vec{w}}
\newcommand{\wvi}{\vec{w}_\text{ideal}}
%matrices
\newcommand{\inv}{^{-1}}
\newcommand{\dg}{^\mathrm{dg}}
\newcommand{\trans}{^\mathrm{T}}
\newcommand{\I}{\mathbf{I}}
%vec of ones
\newcommand{\onev}{\mathbf{e}}
%mat of ones
\newcommand{\onem}{\mathbf{E}}
%Markov matrix
\newcommand{\MM}{\mathbf{Q}}
%prob distributions
\newcommand{\prob}{\mathbf{p}}
\newcommand{\eq}{\prob^\infty}
%first passage times
\newcommand{\fpt}{\mathbf{T}}
%off-diag first passage times
\newcommand{\fptb}{\overline{\fpt}}
%fundamental matrix
\newcommand{\fund}{\mathbf{Z}}
%other symbols for matrices
\newcommand{\Pb}{\mathbf{P}}
\newcommand{\D}{\mathbf{D}}
\newcommand{\pib}{\boldsymbol{\pi}}
\newcommand{\Lb}{\boldsymbol{\Lambda}}
\newcommand{\w}{\mathbf{w}}
\newcommand{\W}{\mathbf{W}}
\newcommand{\M}{\mathbf{M}}
\newcommand{\enc}{\mathbf{q}}
\newcommand{\frg}{\W^{\mathrm{F}}}
\newcommand{\F}{\boldsymbol{\Phi}}
%superscripts
\newcommand{\pot}{^{\text{pot}}}
\newcommand{\dep}{^{\text{dep}}}
\newcommand{\potdep}{^{\text{pot/dep}}}
%sets
\newcommand{\CS}{\mathcal{S}}
\newcommand{\CA}{\mathcal{A}}
\newcommand{\CB}{\mathcal{B}}
\newcommand{\comp}{^\mathrm{c}}
%eigenmodes
\newcommand{\uv}{\mathbf{u}}
\newcommand{\vv}{\mathbf{v}}
\newcommand{\CI}{\mathcal{I}}

%\renewcommand\Fref[1]{}
\providecommand\hyperref[2][1]{#2}

%%%%%%%%%%%%
%% For PNAS Only:
%\url{www.pnas.org/cgi/doi/10.1073/pnas.0709640104}
\renewcommand{\theurl}{www.pnas.org/cgi/doi/10.1073/pnas.0709640104}
\copyrightyear{2016}
\issuedate{Issue Date}
\volume{Volume}
\issuenumber{Issue Number}
%\setcounter{page}{2687} %Set page number here if desired
%%%%%%%%%%%%

@@

\title{Optimal synaptic strategies for different timescales of memory}

\author{Subhaneil Lahiri\affil{1}{Department of Applied Physics, Stanford University, Stanford CA},
\and
Surya Ganguli\affil{1}{}}

\contributor{Submitted to Proceedings of the National Academy of Sciences
of the United States of America}

@@

\begin{article}
\begin{abstract}
  An incredible gulf separates theoretical models of synapses, often described solely by a single scalar value denoting the size of a postsynaptic potential, from the immense complexity of molecular signaling pathways underlying real synapses.
  To understand the functional contribution of such molecular complexity to learning and memory, it is essential to expand our theoretical conception of a synapse from a single scalar to an entire dynamical system with many internal molecular functional states.
  Moreover, theoretical considerations alone demand such an expansion; network models with scalar synapses assuming finite numbers of distinguishable synaptic strengths have strikingly limited memory capacity.
  This raises the fundamental question, how does synaptic complexity give rise to memory?
  To address this, we develop new mathematical theorems elucidating the relationship between the structural organization and memory properties of complex synapses that are themselves molecular networks.
  Moreover, in proving such theorems, we uncover a framework, based on first passage time theory, to impose an order on the internal states of complex synaptic models, thereby simplifying the relationship between synaptic structure and function.

  Overall, we uncover general design principles governing the functional organization of complex molecular networks, and suggest new experimental observables in synaptic physiology, based on first %passage time theory, that connect molecular complexity to memory.
\end{abstract}

\keywords{synaptic plasticity | learning | memory | neuroscience}

\abbreviations{LTP, long term potentiation; 
LTD, long term depression}

\dropcap{I}t is widely thought that our very ability to remember the past over long time scales depends crucially on our ability to modify synapses in our brain in an experience dependent manner.
Classical models of synaptic plasticity model synaptic efficacy as an analog scalar value, denoting the size of a post-synaptic potential injected into one neuron from another.
Theoretical work has shown that such models have a reasonable, extensive memory capacity, in which the number of long term associations that can be stored by a neuron is proportional its number of afferent synapses \cite{Hopfield1982model,Amit1985hopfield,Gardner1988perceptron}.
However, recent experimental work has shown that many synapses are more digital than analog; they cannot robustly assume an infinite continuum of analog values, but rather can only take on a finite number of distinguishable strengths, a number than can be as small as two \cite{Bliss1993LTP,Petersen1998allornone,O'Connor2005switch} (though see \cite{Enoki2009graded}).
This one simple modification leads to a catastrophe in memory capacity:  classical models with digital synapses, when operating in a palimpsest mode in which the ongoing storage of new memories can overwrite previous memories, have a memory capacity proportional to the logarithm of the number of synapses \cite{amit1992constraints,amit1994learning}.
Intuitively, when synapses are digital, the storage of a new memory can flip a population of synaptic switches, thereby rapidly erasing previous memories stored in the same synaptic population. This result indicates that the dominant theoretical basis for the storage of long term memories in modifiable synaptic switches is flawed.

@@

Theoretical studies of complex, metaplastic synapses have focused on analyzing the memory performance of a limited number of very specific molecular dynamical systems, characterized by a number of internal states in which potentiation and depression each induce a specific set of allowable transitions between states (\eg see \Fref{fig:models} below).
While these models can vastly outperform simple binary synaptic switches, these analyses leave open several deep and important questions.
For example, how does the structure of a synaptic dynamical system determine its memory performance?  What are the fundamental limits of memory performance over the space of all possible synaptic dynamical systems?
What is the structural organization of synaptic dynamical systems that achieve these limits?
Moreover, from an experimental perspective, it is unlikely that all synapses can be described by a single canonical synaptic model;  just like the case of neurons, there is an incredible diversity of molecular networks underlying synapses both across species and across brain regions within a single organism \cite{Emes2012synapserev}.
In order to elucidate the functional contribution of this diverse molecular complexity to learning and memory, it is essential to move beyond the analysis of specific models and instead develop a general theory of learning and memory for complex synapses.
Moreover, such a general theory of complex synapses could aid in development of novel artificial memory storage devices.  % Sutskever, memory for learning?

@@

Much of the previous work on these types of complex synaptic models has focused on understanding the memory curves of specific models, or choices of $\M\potdep$.
Two examples of these models are shown in \Fref{fig:models}. We see that they have different memory properties.
The serial model performs relatively well at one particular timescale, but it performs poorly at other times.
The cascade model does not perform quite as well at that time, but it maintains its performance over a wider range of timescales.

\begin{figure}[tbp]
 \begin{center}
 \begin{myenuma}
  \item\aligntop{\includegraphics[width=0.25\linewidth]{cascade.svg}}\label{fig:cascade}
  \item\aligntop{\includegraphics[width=0.25\linewidth]{multistate.svg}}\label{fig:multistate}
  \item\aligntop{\includegraphics[width=0.3\linewidth]{cascms.eps}}\label{fig:cascms}
 \end{myenuma}
 \end{center}
  \caption{Models of complex synapses.
  (\ref{fig:cascade}) The cascade model of \cite{Fusi2005cascade}, showing transitions between states of high/low synaptic weight (red/blue circles) due to potentiation/depression (solid red/dashed blue arrows).
  (\ref{fig:multistate}) The serial model of \cite{Leibold2008serial}.
  (\ref{fig:cascms}) The memory curves of these two models, showing the decay of the signal-to-noise ratio (to be defined \hyperref[sec:setup]{below}) as subsequent memories are stored.
  }\label{fig:models}
\end{figure}

@@

In the previous section, in \eqref{eq:SNRcalc} we have described an analytic expression for a memory curve as a function of the structure of a synaptic dynamical system, described by the pair of stochastic transition matrices  $\M\potdep$.
Since the performance measure for memory is an entire memory curve, and not just a single number, there is no universal scalar notion of optimal memory in the space of synaptic dynamical systems.
Instead there are tradeoffs between storing proximal and distal memories;  often attempts to increase memory at late (early) times by changing $\M\potdep$, incurs a performance loss in memory at early (late) times in specific models considered so far \cite{Fusi2005cascade,Fusi2007multistate,Leibold2008serial}.
Thus our end goal, achieved in \sref{sec:env}, is to derive an envelope memory curve in the SNR-time plane, or a curve that forms an upper-bound on the {\it entire} memory curve of {\it any} model.
In order to achieve this goal, in this section, we must first derive upper bounds, over the space of all possible synaptic models, on two different scalar functions of the memory curve: its initial SNR, and the area under the memory curve.
In the process of upper-bounding the area, we will develop an essential framework to organize the structure of synaptic dynamical systems based on first passage time theory.

@@

To maximize this flux, potentiation from a weak state must be guaranteed to end in a strong state, and depression must do the reverse.
An example of such a model is shown in \Fref{fig:max_init}(\ref{fig:max_init_pot},\ref{fig:max_init_dep}).
These models have a property known as ``lumpability'' (see \cite[\S6.3]{kemeny1960finite} for the discrete time version and \cite{burke1958markovian,Ball1993Lumpability} for continuous time).
They are completely equivalent (\ie have the same memory curve) as a two state model with transition probabilities equal to 1, as shown in \Fref{fig:max_init}(\ref{fig:binary_det}).

@@

With the states in the order described above, we can find perturbations of $\M\potdep$ that will always increase the area, whilst leaving the equilibrium distribution, $\eq$, unchanged.
Some of these perturbations are shown in \Fref{fig:perts}, see supplementary material for details.
For example, in \Fref{fig:perts}(\ref{fig:tri_right}), for two states $i$ on the left and $j$ on the right, with $j$ being more ``potentiated" than $i$ (i.e. $\smash{\eta^+_i > \eta^+_j}$), we have proven that increasing $\smash{\M\pot_{ij}}$ and decreasing $\smash{\M\dep_{ij}}$ leads to an increase in area.
The only thing that can prevent these perturbations from increasing the area is when they require the decrease of a matrix element that has already been set to 0.
This determines the topology (non-zero transition probabilities) of the model with maximal area.
It is of the form shown in \Fref{fig:env}(\ref{fig:sticky}),%\Fref{fig:max_area},
with potentiation moving one step to the right and depression moving one step to the left.
Any other topology would allow some class of perturbations (\eg\ in \Fref{fig:perts}) to further increase the area.

\begin{figure}[tbp]
 \begin{center}
 \begin{myenuma}
  \item\aligntop{\includegraphics[width=0.14\linewidth]{triangular_right.svg}}\label{fig:tri_right}\hp
  \item\aligntop{\includegraphics[width=0.14\linewidth]{triangular_left.svg}}\label{fig:tri_left}\hp
  \item\aligntop{\includegraphics[width=0.28\linewidth]{shortcut.svg}}\label{fig:shortcut}
 \end{myenuma}
 \end{center}
  \caption{Perturbations that increase the area.
  (\ref{fig:tri_right}) Perturbations that increase elements of $\M\pot$ above the diagonal and decrease the corresponding elements of $\M\dep$. It can no longer be used when $\M\dep$ is lower triangular, \ie depression must move synapses to ``more depressed'' states.
  (\ref{fig:tri_left}) Perturbations that decrease elements of $\M\pot$ below the diagonal and increase the corresponding elements of $\M\dep$. It can no longer be used when $\M\pot$ is upper triangular, \ie potentiation must move synapses to ``more potentiated'' states.
  (\ref{fig:shortcut}) Perturbation that decreases ``shortcut'' transitions and increases the bypassed ``direct'' transitions. It can no longer be used when there are only nearest-neighbor ``direct'' transitions.
  }\label{fig:perts}
\end{figure}

As these perturbations do not change the equilibrium distribution, this means that the area of \emph{any} model is bounded by that of a linear chain with the same equilibrium distribution.
The area of a linear chain model can be expressed directly in terms of its equilibrium state distribution, $\eq$, yielding the following upper bound on the area of any model with the same $\eq$ (see supplementary material):
%
\begin{equation}\label{eq:multistatearea}
  A \leq \frac{2\sqrt{N}}{r} \sum_k \brk{k - \sum_j j\eq_j} \eq_k \w_k
    = \frac{2\sqrt{N}}{r} \sum_k \abs{k - \sum_j j\eq_j} \eq_k ,
\end{equation}
%
where we chose $\w_k=\sgn[{k - \sum_j j\eq_j}]$.
We can then maximize this by pushing all of the equilibrium distribution symmetrically to the two end states.
This can be done by reducing the transition probabilities out of these states, as in \Fref{fig:env}(\ref{fig:sticky}). %\Fref{fig:max_area}.
This makes it very difficult to exit these states once they have been entered.
The resulting area is
%
\begin{equation}\label{eq:max_area}
  A \leq \frac{\sqrt{N}(M-1)}{r}.
\end{equation}
%
This analytical result is similar to a numerical result found in \cite{Barrett2008discrete} under a slightly different information theoretic measure of memory performance.

@@

Now we can pick some fixed time, $\smash{t_0}$, and maximize the SNR at that time \wrt the parameters $\smash{\brc{\CI_a,\tau_a}}$, subject to the constraints above.
This always results in a single nonzero $\smash{\CI_a}$; in essence, optimizing memory at a single time requires a single exponential.  The resulting optimal memory curve, along with the achieved memory at the chosen time, depends on $t_0$ as follows:
%
\begin{equation}\label{eq:env}
\begin{aligned}
  t_0 &\leq \frac{M-1}{r} &
   \implies
  \snr(t) &= \sqrt{N}\e^{-rt/(M-1)} &
   \implies
  \snr(t_0) &= \sqrt{N}\e^{-rt_0/(M-1)},\\
  t_0 &\geq \frac{M-1}{r} &
   \implies
  \snr(t) &= \frac{\sqrt{N}(M-1)\e^{-t/t_0}}{rt_0} &
   \implies
  \snr(t_0) &= \frac{\sqrt{N}(M-1)}{\e rt_0}.
\end{aligned}
\end{equation}
%
Both the initial SNR bound and the area bound are saturated at early times.
At late times, only the area bound is saturated.
The function $\snr(t_0)$, the green curve in \Fref{fig:env}(\ref{fig:envplot}), above forms a memory curve envelope with late-time power-law decay $\sim \smash{t_0\inv}$.
No synaptic model can have an SNR that is greater than this at any time.
We can use this to find an upper bound on the memory lifetime, $\tau(\epsilon)$, by finding the point at which the envelope crosses $\epsilon$:
%
\begin{equation}\label{eq:lifebound}
  \tau(\epsilon) \leq \frac{\sqrt{N}(M-1)}{\epsilon\e r},
\end{equation}
%
where we assume $N>(\epsilon\e)^2$.
Intriguingly, both the lifetime and memory envelope expand linearly with the number of internal states $M$, and increase as the square root of the number of synapses $N$.

This leaves the question of whether this bound is achievable.
At any time, can we find a model whose memory curve touches the envelope?
The red curves in \Fref{fig:env}(\ref{fig:envplot}) show the closest we have come to the envelope with actual models, by repeated numerical optimization of $\SNR(t_0)$ over $\M\potdep$ with random initialization and by hand designed models.

We see that at early, but not late times, there is a gap between the upper bound that we can prove and what we can achieve with actual models.
There may be other models we haven't found that could beat the ones we have, and come closer to our proven envelope.
However, we suspect that the area constraint is not the bottleneck for optimizing memory at times less than  $\CO(\frac{M}{r})$.
We believe there is some other constraint that prevents models from approaching the envelope, and currently are exploring several mathematical conjectures for the precise form of this constraint in order to obtain a potentially tighter envelope.
Nevertheless, we have proven rigorously that no model's memory curve can ever exceed this envelope, and that it is at least tight for late times, longer than $\CO(\frac{M}{r})$, where models of the form in \Fref{fig:env}(\ref{fig:sticky})%\Fref{fig:max_area}
 can come close to the envelope.

\begin{figure}
 \begin{center}
 \parbox[t]{0.44\linewidth}{
 \begin{myenuma}
  \item\aligntop{\includegraphics[width=0.95\linewidth]{env.svg}}\label{fig:envplot}
 \end{myenuma}
  }\hp
  \parbox[t]{0.45\linewidth}{
 \begin{myenuma}
 \addtocounter{myenumai}{1}
  \item\aligntop{\includegraphics[width=0.8\linewidth]{diffjump.svg}}\label{fig:diffjump}\\[0.1\linewidth]
  \item\aligntop{\includegraphics[width=0.8\linewidth]{multistate_sticky.svg}}\label{fig:sticky}
 \end{myenuma}
  }
 \end{center}
  \caption{The memory curve envelope for $N=100$, $M=12$.
  (\ref{fig:envplot}) An upper bound on the SNR at any time is shown in green.
  The red dashed curve shows the result of numerical optimization of synaptic models with random initialization.
  The solid red curve shows the highest SNR we have found with hand designed models.
  At early times these models are of the form shown in (\ref{fig:diffjump}) with different numbers of states, and all transition probabilities equal to $1$.
  At late times they are of the form shown in (\ref{fig:sticky}) with different values of $\varepsilon$.
  The model shown in (\ref{fig:sticky}) also saturates the area bound \eqref{eq:max_area} in the limit $\varepsilon\to0$.
  }\label{fig:env}
\end{figure}
